---
title: "Applied R for Social Scientists"
author: "J. Alexander Branham"
date: "April 2016"
header-includes: 
    - \usetheme[titleformat=smallcaps, progressbar=frametitle]{metropolis}
output: 
  beamer_presentation:
    latex_engine: xelatex
classoption: aspectratio=169
---

# This Class
## This class
* Assumption of *some* previous exposure to R
  + We're not explaining assignment, packages, calling functions, etc
* Will *not* be covering statistics
* Alex will start with some common data tasks: loading, variable
  creation, merging, etc
* Daniel will take the second part to talk about tables and
  visualization 

# Getting the data
## Download the data

* `source()` runs a file through R
* This one checks if you have the data already and tries to download
  it if not
* The dataset we're using is the General Social Survey spanning 1972-2014

```{r}
source("check-gss-and-maybe-download.R")
```

## Reading data
* R can read almost any data
* Here are some of the most common types:

| package | function   | file formats               |
|---------+------------+----------------------------|
| foreign | `read.*`   | dta, spss, etc             |
| haven   | `read_*`   | dta (13+) files and others |
| readr   | `read_csv` | csv files                  |
  
## Read data
* We have stata data (`*.dta`)
* `convert.factors = FALSE` ensures that R doesn't convert the values to
  the labels that stata uses

```{r}
GSS <- foreign::read.dta("./data/GSS7214_R5.DTA",
                         convert.factors = FALSE)
```

# Data management
## The data: a table
* Let's get a sense of the data we're working with
* Do you want more, less, or about the same spending? (education and
  social security)

```{r}
table(GSS$nateduc, GSS$natsoc, exclude = NULL)
```

## correlation
* `cor` gives us (by default) the pearson's $r$ between two variables
* Without setting `use`, R tries to use the whole data, some of which
  are missing and thus results in `NA`

```{r}
cor(GSS$nateduc,
    GSS$natsoc,
    use = "complete.obs")
```

## Variable creation 
* Let's make an indicator variable for whether a respondent is black
  or not
* Here's the `race` variable
* I also like to make sure that I'm not going to overwrite an existing
  variable
 

```{r}
table(GSS$race)
```

```{r}
table(GSS$black)
```

## Variable creation
* Using `ifelse` to create a variable conditional on other var's
  values

```{r}
GSS$black <- ifelse(GSS$race == 2, TRUE, FALSE)
table(GSS$black)
```

## Subsetting observations
* Now let's check to see if that correlation is different for black
  people
* Note how ugly this looks!

```{r}
cor(GSS$nateduc[GSS$black == TRUE],
    GSS$natsoc[GSS$black == TRUE],
    use="complete.obs")
```

## dplyr
* `dplyr` is an R package that makes data management *much* easier
* Different functions for data munging:
  - `filter()`, `select()`, `mutate()`
* It introduces the pipe operator `%>%` to the language
* Functions for merging data
  - `*_join`: full, inner, left, right
* `group_by`, which lets us perform operations on groups of the data
* Because I'll use `tidyr` later and it gets angry if you load it
  after `dplyr`, I'm loading it now

```{r}
library(tidyr)
suppressPackageStartupMessages(library(dplyr))
```

## Subsetting the dplyr way
* The pipe (`%>%`) "pipes" the output of the last thing into the first
  argument of the next thing
* `summarize` (or `summarise`) from `dplyr` returns a `data.frame`

```{r}
with(filter(GSS, black == TRUE),
     cor(nateduc, natsoc,
         use = "complete.obs"))
```

## Subsetting the dplyr way 
```{r}
GSS %>%
  filter(black == TRUE) %>%
  summarize(mycor =
              cor(nateduc, natsoc,
                  use = "complete.obs"))
```

## Dropping observations
* 1972 doesn't have any observations we're interested in (our spending
  variables weren't asked), so let's drop it
* Again, we can use `filter`, but this time we assign the result back
  to `GSS`:

```{r}
GSS <- GSS %>%
  filter(year != 1972)
```

## Factors
* Variables with categories can be represented as factors in R
* If you want R to think they're ordered, you can use `ordered ` TRUE=
  as an argument

 <!-- GSS$sex <- factor(GSS$sex, -->
 <!--                   levels=c(1, 2), -->
 <!--                   labels=c("M", "F")) -->

```{r}
table(GSS$sex)
```

## Factors 
```{r}
GSS <- GSS %>%
  mutate(sex = factor(sex,
                      levels = c(1, 2),
                      labels = c("M", "F")))

table(GSS$sex)
```

## Operations by subcategories
* `dplyr` provides `group_by`
* Lets us perform operations to grouped data


```{r}
thecors <- GSS %>%
  group_by(sex, black) %>%
  summarize(thecor = cor(nateduc, natsoc,
                         use = "complete.obs"),
            n = n())
```

## Operations by subcategories

```{r}
print(thecors)
```

## Aggregation 
* Maybe we're interested in preferences by year?

```{r}
gss_yearly <- GSS %>%
  group_by(year) %>%
  summarize(educ = mean(nateduc,
                        na.rm = TRUE),
            soc = mean(natsoc,
                       na.rm = TRUE))
```

## Aggregation

```{r}
head(gss_yearly)
```

## Functions
* Means are nice, but there are other ways to summarize data
* What if we want to look at the proportion of people who support more
  spending minus the proportion who support less? 


```{r}
netsupport <- function(thedata){
  prop_more <- mean(thedata == 1, na.rm = TRUE)
  prop_less <- mean(thedata == 3, na.rm = TRUE)
  prop_more - prop_less
}
``` 

## Functions

```{r}
GSS %>%
  group_by(year) %>%
  summarize(support_educ = netsupport(nateduc),
            support_soc = netsupport(natsoc))
```

## Merging 
* The `ggplot2` package provides the `economics` data.frame that has
  US economic data starting in July 1967
* `?economics` gives more info 

```{r}
library(ggplot2)
head(economics, 3)
```

## Merging
* Let's make an unemployment rate by unemploy*pop

```{r}
economics <- economics %>%
  mutate(unemp_rate = unemploy / pop)
```

* Note `mutate` is from dplyr, this is base R:

```{r}
economics$unemp_rate <- economics$unemploy / economics$pop
```


## Merging
* The `economics` data is monthly and our GSS data is yearly, so we
  need to aggregate 

```{r}
economics_yearly <- economics %>%
  mutate(year = format(date, "%Y")) %>%
  group_by(year) %>%
  summarize(unemp = mean(unemp_rate))
```

## Merging
* Let's see what our data looks like now!

```{r}
head(economics_yearly)
```

## Merging
* Now we have two data.frame objects --- `gss_yearly` and
  `economics_yearly` --- that we want to join together
* `dplyr` provides a really easy way of doing this
* The jargon comes from SQL, a programming language used to store data
* What you probably call a "merge" dplyr calls a "join"
* `*_join` where `*` is either `full`, `inner`, `left`, or `right`
* We'll use `left_join` since the economics data contains years that
  aren't in the GSS

## Merging

```{r, error = TRUE}
gss_yearly <- left_join(gss_yearly,
                        economics_yearly,
                        by = "year")
```

## Errors
* Error: cannot join on columns 'year' x 'year': Can't join on 'year'
  x 'year' because of incompatible types (character / integer)

## Errors
* The error on the last slide indicates that the `year` variable in
  the two datasets is different
* Let's verify that:

```{r}
class(gss_yearly$year)
```

```{r}
class(economics_yearly$year)
```


## Merging
* Solution: change `economics_yearly$year` to an integer

```{r}
economics_yearly$year <- as.integer(economics_yearly$year)

gss_yearly <- left_join(gss_yearly,
                        economics_yearly,
                        by="year")
```

## Merging 

```{r}
head(gss_yearly)
```

## Writing data
* Maybe you want to save this new data so you don't have to re-run the
  merging whenever you want to 

| package | function     | result     |
|---------+--------------+------------|
| readr   | `write_csv`  | csv file   |
| utils   | `write.csv`  | csv file   |
| base    | `save`       | Rdata file |
| xlsx    | `write.xlsx` | excel file |

* R can also write to stata/SPSS/SAS files through `foreign` or `haven`

## Writing data
* Let's save a csv file
* If the `data/` subfolder doesn't exist, this will produce an error
* The script that we ran at the beginning created this if it didn't
  already exist


```{r}
readr::write_csv(gss_yearly, "data/gss-yearly-data.csv")
```

# Tidying data
## What is tidy data?
* Sometimes the data you get aren't *tidy*
* Tidy data are data where each row is an observation, each column a
  variable, and each cell a value
* Most of the strategies I showed you above assume that you're dealing
  with tidy data
* Remember I loaded tidyr earlier, so there's no need to call
  `library` again

## Example of untidy data

```{r}
messy1 <- data_frame(
  country = c("Afghanistan", "Albania", "Algeria"), 
  "2007" = c(43.82, 76.42, 72.30), 
  "2002" = c(42.13, 75.65, 70.99))
```

## Untidy data

```{r}
print(messy1)
```

## Use gather when you have non-variable columns
* `gather` can also turn wide to long

```{r}
gather(messy1, "year", "life_expect", 2:3)
```

## Untidy data

```{r}
messy2 <- data.frame(
  country = c(rep("Afghanistan", 4), rep("Albania", 4), rep("Algeria", 4)), 
  year = c(rep(2002, 2), rep(2007, 2)), 
  variable = c("life_expect", "pop"), 
  value = c(42.12, 25268405, 43.82, 31889923,
            75.65, 3508512, 76.42, 3600523,
            70.99, 31287142, 72.30, 33333216)
)
```

## Untidy data

```{r}
head(messy2)
```

## use spread!
* `spread` can also turn long to wide

```{r}
spread(messy2, key = variable, value)
```

## separate and unite
* If you have two variables in one column, use `separate`
  * For example, a rate of # of people with a trait / total population
    in each country
* One variable across two columns? use `unite`
  * one column for century (19, 20) and another for year (00... 09)

# Iteration: Loops and apply

## DRY
* DRY (*Don't Repeat Yourself*) is an acronym from computer science
* Repeating yourself makes your code harder to deal with:
  * Intent is less clear
  * Harder to spot bugs

## Loops
* For loops get lots of hate online because people think they're slow
  (they aren't)
* They can be hard to read, though


```{r}
thedata <- data_frame(
  one = rnorm(100), two = rnorm(100), 
  three = rnorm(100), four = rnorm(100)
)
```


## For loops

```{r}
output <- list()
output[[1]] <- median(thedata$one)
output[[2]] <- median(thedata$two)
output[[3]] <- median(thedata$three)
output[[4]] <- median(thedata$four)
print(output); rm(output)
```

## For loops
```{r}
output <- list()
for (i in 1:4) {
  output[[i]] <- median(thedata[[i]])
}
print(output); rm(output)
```

## Mapping functions
* Of course, we oftentimes need to perform an operation across many
  columns
* This is where the `map` family (from `purrr`) steps in:

```{r}
library(purrr)
map(thedata, median)
```

## Mapping functions

```{r}
map_dbl(thedata, median)
```

# Nested data 
## For fun: nested date
* Some data is nested in a hierarchical way
* the `gapminder` data are a good example[^1]

```{r}
library(gapminder); library(ggplot2)
head(gapminder, 3)
```

## Life expectancy over time


```{r, eval = FALSE} 
ggplot(gapminder, aes(x = year, y = lifeExp,
                      color = continent, by = country)) +
  geom_line()
```

## life expectancy over time
```{r, echo = FALSE} 
ggplot(gapminder, aes(x = year, y = lifeExp,
                      color = continent, by = country)) +
  geom_line()
```


## introducing the nest function

```{r}
by_country <- gapminder %>%
  group_by(continent, country) %>%
  nest()
```

## nest
* Now we have a data frame with one row per group and a column where
  each cell is itself a whole data frame

```{r}
head(by_country,3)
```

## nest
* So for example the first element of the `data` column contains the
  whole data frame for Afghanistan 

```{r}
by_country$data[[1]]
```

## nest
* You can create a linear model for each country then:


```{r}
by_country <- by_country %>%
  mutate(model = map(data, 
                     ~ lm(lifeExp ~ year, data = .)))
```

## nest

```{r}
head(by_country, 3)
```

## nest
* Here we can extract the fitted values and plot a lint of the fitted values
* By continent, country 

```{r, eval = FALSE}
by_country %>% unnest(model %>% map(broom::augment)) %>%
  select(continent, country, year, .fitted) %>%
  ggplot(aes(x = year, y = .fitted,
             by = country, color = continent)) +
  geom_line()
```


## nest 
```{r, echo = FALSE}
by_country %>% unnest(model %>% map(broom::augment)) %>%
  select(continent, country, year, .fitted) %>%
  ggplot(aes(x = year, y = .fitted,
             by = country, color = continent)) +
  geom_line()
```

[^1]: This example taken from the [blog post](http://blog.rstudio.org/2016/02/02/tidyr-0-4-0/)
